---
title: "Highlighting high influence observations in ggplot2"
author: "Individual Study"
date: "10/12/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

The next task is to use 'base ggplot' to show high influence observations.

Below, you will see two plots produced with 'base R graphics'.  These are from - https://www.stat.berkeley.edu/~spector/s133/Lr1.html 

You will reproduce with ggplot2.

## First we read in data, and limit to variables of interest, Alcohol and Proline.

```{r}
library(tidyverse)
wine_raw <- read_csv("https://gist.githubusercontent.com/tijptjik/9408623/raw/b237fa5848349a14a14e5d4107dc7897c21951f5/wine.csv")

head(wine_raw)

wine_raw %>% 
  select(Alcohol, Proline) ->
wine

head(wine)

```


## Then we estimate a linear model (ols).

```{r}
# estimate a linear model and assign to simple.lm
# we will only be looking at simple lm (one predictor) for the independent study
simple.lm <- lm(formula = Alcohol ~ Proline,
                data = wine_raw)
```

## Once estimated, we can get a vector of cooks distances and leverage values for each observation.  We add these to the data frame.

```{r}
# save cooks distance for each observation using $ notation
# Cook, R. Dennis (February 1977). "Detection of Influential Observations in Linear Regression". Technometrics. American Statistical Association. 19 (1): 15–18. doi:10.2307/1268249. JSTOR 1268249. MR 0436478.
# https://en.wikipedia.org/wiki/Cook%27s_distance
# as another variable in the wine data frame
wine$cooks <- cooks.distance(model = simple.lm)

# save 'hat' calculation (leverage I think - we'll have to come back to exact calculation)
# as another variable inthe wine data frame
wine$hat <-  lm.influence(model = simple.lm)$hat
wine
```

## What follows are the examples from https://www.stat.berkeley.edu/~spector/s133/Lr1.html which build plots using base r graphics.

### Note: looking just at the code `wine$cooks > quantile(wine$cooks, .90)` might be helpful.  What does `quantile(wine$cooks, .90)` return?  And the inequality?

```{r}
# base R plot highlighting 10% highest cooks
plot(x = wine$Proline, y = wine$Alcohol,
     col = ifelse(test = wine$cooks >
                    quantile(wine$cooks, .90),
                  yes = 'red',
                  no = 'black'))

# base R plot highlighting 10% highest leverage (hat)
plot(x = wine$Proline,
     y = wine$Alcohol,
     col=ifelse(wine$hat > quantile(wine$hat,.90),
                'blue','black'))
```

## Now, with ggplot2, we'll use two point layers to build very similar plots

- First, plot all the observations. 
- Then, plot 10% of observations with largest cook's (or largest hat) in highlight color.

### Need a hint?  Below, the second point layer highlights a subset of points.

### Note: in base R we refer to variables w/ dollar sign notation, but in tidyverse/ggplot2, you can use variable names directly (i.e. Proline instead of wine$Proline)

```{r}
# An example of defining subset of data for a point layer
ggplot(data = wine) + 
  aes(x = Proline) + 
  aes(y = Alcohol) + 
  geom_point() + # this point layer ALL data default color black
  geom_point(data = wine %>% # this next point layer
               filter(Proline > 500), # SUBSET of data
             color = "green") # custom color
```

# Task 4.2

Now, Instead of highlighting a subset, label each point with cooks distance.  (hint - you can use the geom_text()).

Here is an example of where the observations are labeled with the Alcohol level. 

What does nudge_y do? What does check_overlap do?

```{r}
ggplot(data = wine) + 
  aes(x = Proline) + 
  aes(y = Alcohol) + 
  geom_point(color = "darkgrey") + 
  aes(label = round(Alcohol, 1)) +
  geom_text(nudge_y = .1, 
            check_overlap = T)
```


Right now we aren't writing any new functions, rather just using ggplot2 'out-of-the-box'.  



## Task 4. Model diognostics (maybe a bit more together)...

- https://joshualoftus.com/posts/2020-11-23-least-squares-as-springs/
- https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Lecture/lecture08_2020JC.html#7




